---
title: "PleioMatchR: from 'significant but direction-random' to stratified diagnosis"
author: "PleioMatchR developers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PleioMatchR workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
set.seed(20260908)
```

## Motivation

A recurring finding in cross-trait GWAS analyses is *"enrichment is
significant but the direction of effects looks random"*. This pattern can
reflect genuine polygenic noise, but it can also hide two (or more)
biologically distinct subgroups whose effects are internally consistent yet
point in opposite directions. PleioMatchR is a matched-control diagnostic
toolbox designed to tell these stories apart.

This vignette reproduces the full workflow on the simulated
T2D-to-breast-cancer style dataset shipped with the package.

```{r setup}
library(PleioMatchR)
data(demo_sentinels, package = "PleioMatchR")
data(demo_outcome, package = "PleioMatchR")
data(demo_ld_scores, package = "PleioMatchR")

head(demo_sentinels)
table(demo_sentinels$pathway)
```

## Step 1 - import and harmonise

`import_sumstats()` recognises common GWAS column aliases and returns a
canonical `pmr_sumstats` table:

```{r import}
exp <- import_sumstats(demo_sentinels)
out <- import_sumstats(demo_outcome)
class(exp)
head(exp)
```

If exposure and outcome tables contain overlapping variants measured in
different allele orientations, align them first with
`harmonise_sumstats()`:

```{r harmonise, eval = FALSE}
h <- harmonise_sumstats(exp, out)
h$report
```

## Step 2 - genome-aware control pool

The manuscript-style matching defaults are: same chromosome, MAF within
±0.03, physical distance > 500 kb from each sentinel, and exclusion of
outcome genome-wide-significant variants. Supplying LD scores adds an
LD-score stratum constraint so that cases and controls share a similar
genomic background:

```{r controls}
ctrl_plain <- build_control_pool(demo_sentinels, demo_outcome,
                                 n_controls = 10, seed = 42)
ctrl_lds <- build_control_pool(demo_sentinels, demo_outcome,
                               n_controls = 10,
                               ld_scores = demo_ld_scores, seed = 42)

ctrl_lds
summary(ctrl_lds)
```

Compare matching quality:

```{r balance}
cbind(
  plain = ctrl_plain$diagnostics$balance,
  lds   = ctrl_lds$diagnostics$balance
)
```

## Step 3 - enrichment and direction test

```{r test}
res <- pleio_test(demo_sentinels, demo_outcome, controls = ctrl_lds,
                  n_perm = 10000, seed = 1)
res
```

The enrichment index (EI = observed/expected hits) is highly significant.
The direction test, however, finds 15 positive / 15 negative effects:
perfectly balanced and statistically indistinguishable from random sign
flipping (p = 1). WDC ≈ 0 and I² ≈ 97% quantify the directional chaos.

```{r plots, fig.show = 'hold'}
plot(res, which = "butterfly")
plot(res, which = "null")
plot(res, which = "forest")
```

The butterfly plot carries a heuristic PRS-suitability flag: despite strong
enrichment, this group of loci is **not** suitable for straightforward
outcome PRS construction without further modelling.

## Step 4 - is the randomness real or stratified?

Both the obesity axis and the islet axis contain 15 sentinels. Re-running the
test inside each axis changes the picture completely:

```{r stratify}
annot <- data.frame(snp = demo_sentinels$snp,
                    pathway = demo_sentinels$pathway)
st <- stratify_by_pathway(res, annot, n_perm = 5000, seed = 1)
st
```

Obesity-axis loci are 15/15 positive and islet-axis loci 15/15 negative.
This is exactly the *"shared mechanism, opposite direction by subtype"*
scenario: the correct scientific conclusion is not "no sharing" but "two
consistent, opposing pathways".

## Step 5 - cross-ancestry drift check

If the exposure is measured in one ancestry and the outcome in another,
check allele-frequency drift before trusting direction tests:

```{r ancestry}
drift <- cross_ancestry_check(
  demo_sentinels[1:20, c("snp", "ea", "oa", "eaf")],
  demo_sentinels[1:20, c("snp", "ea", "oa", "eaf")],
  label_x = "EAS", label_y = "EUR"
)
drift
```

Real usage supplies the two ancestry panels' allele frequencies for the same
variants (e.g. extracted from 1000 Genomes); the function flags palindromic
variants with non-informative MAF and variants with large |dAF| as candidates
for exclusion or strand verification.

## Calibration: what the simulations say

The permutation test is exact conditional on each matched set, but it is
good practice to confirm type-I error behaviour empirically. The following
loop simulates 30 null datasets (no enrichment) and 10 enriched datasets:

```{r calibration}
run_null_sim <- function(seed, n_sent = 12L, n_perm = 200L) {
  set.seed(seed)
  pos_sent <- seq(2e6, by = 3e6, length.out = n_sent)
  pos_ctrl <- as.vector(vapply(pos_sent, function(pp)
    pp + c(0.7, 1.0, 1.5, 2.0, 2.5, 2.7) * 1e6, numeric(6)))
  rows <- rbind(
    data.frame(snp = paste0("s", seq_len(n_sent)), chr = 1L,
               pos = pos_sent, stringsAsFactors = FALSE),
    data.frame(snp = paste0("d", seq_along(pos_ctrl)), chr = 1L,
               pos = pos_ctrl, stringsAsFactors = FALSE))
  rows$ea <- sample(c("A", "C", "G", "T"), nrow(rows), replace = TRUE)
  rows$oa <- sample(c("A", "C", "G", "T"), nrow(rows), replace = TRUE)
  rows$eaf <- runif(nrow(rows), 0.1, 0.9)
  rows$se <- 0.035
  rows$p <- runif(nrow(rows), 0, 0.06)
  sent <- data.frame(snp = paste0("s", seq_len(n_sent)), chr = 1L,
                     pos = pos_sent, ea = rows$ea[seq_len(n_sent)],
                     oa = rows$oa[seq_len(n_sent)],
                     eaf = rows$eaf[seq_len(n_sent)],
                     beta = 0.1, se = 0.03, p = 1e-10,
                     stringsAsFactors = FALSE)
  ctrl <- build_control_pool(sent, rows, n_controls = 6L, maf_tol = 0.5,
                             seed = seed, verbose = FALSE)
  pleio_test(sent, rows, controls = ctrl, p_threshold = 0.03,
             window_kb = 100, n_perm = n_perm, seed = seed,
             verbose = FALSE)$enrichment$p_upper
}

p_null <- vapply(1:30, run_null_sim, numeric(1))
p_pow  <- vapply(31:40, function(s) {
  set.seed(s)
  ## same layout as run_null_sim() but with all case rows below threshold
  pos_sent <- seq(2e6, by = 3e6, length.out = 12L)
  pos_ctrl <- as.vector(vapply(pos_sent, function(pp)
    pp + c(0.7, 1.0, 1.5, 2.0, 2.5, 2.7) * 1e6, numeric(6)))
  rows <- rbind(data.frame(snp = paste0("s", 1:12), chr = 1L,
                           pos = pos_sent, stringsAsFactors = FALSE),
                data.frame(snp = paste0("d", 1:72), chr = 1L,
                           pos = pos_ctrl, stringsAsFactors = FALSE))
  rows$ea <- sample(c("A", "C", "G", "T"), nrow(rows), replace = TRUE)
  rows$oa <- sample(c("A", "C", "G", "T"), nrow(rows), replace = TRUE)
  rows$eaf <- runif(nrow(rows), 0.1, 0.9)
  rows$se <- 0.035
  rows$p <- runif(nrow(rows), 0, 0.06)
  rows$p[1:12] <- 1e-9
  sent <- data.frame(snp = paste0("s", 1:12), chr = 1L, pos = pos_sent,
                     ea = rows$ea[1:12], oa = rows$oa[1:12],
                     eaf = rows$eaf[1:12], beta = 0.1, se = 0.03,
                     p = 1e-10, stringsAsFactors = FALSE)
  ctrl <- build_control_pool(sent, rows, n_controls = 6L, maf_tol = 0.5,
                             seed = s, verbose = FALSE)
  pleio_test(sent, rows, controls = ctrl, p_threshold = 0.03,
             window_kb = 100, n_perm = 200L, seed = s,
             verbose = FALSE)$enrichment$p_upper
}, numeric(1))

cat("Type-I (null): mean p =", mean(p_null),
    "| p<0.05:", sum(p_null < 0.05), "/ 30\n")
cat("Power: p<0.05:", sum(p_pow < 0.05), "/ 10\n")
```

Null p-values are approximately uniform (expected ~1.5 of 30 below 0.05),
and the enriched scenario is detected in nearly every replicate. The same
loop, scaled up (thousands of replicates, realistic LD, multiple ancestry
scenarios, mixed directions), is the required simulation appendix for a
methods paper built on this package.

## Where to go next

* `?pleio_test`, `?build_control_pool`, `?stratify_by_pathway`
* `?direction_heterogeneity` (WDC/I2, still descriptive)
* `?cross_ancestry_check` (diagnostic only)
